{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2023-06-13 Docs/manuscript.Rnw
edited intro/discussion

% emwJun4: I really love the info you added here, but we have to be careful to not focus too deeply on synchrony as it is a very specific topic and our work is broader than that (sorry if I confused you by my request for 'refs' -- I mostly wanted just citations; I suspect we can use some of the text for the discussion so it will not go to waste). We also need to keep paragraphs themselves and our intro overall short, so I cut here. Thus I slimmed this way down and kept many references, but less of the examples.
Many of the currently observed responses to climate change appear to be mostly plastic \citep{burton22,zettle21,bonamour19, king17}, including the most reported biological response to climate change---shifting phenology. Phenology (the timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and exploit the resources of each growing season \citep{chuine10,hanninen11,rytteri21,posle18}. As such, phenology plays a significant role in determining fitness for both plants \citep{guo22,chuine01} and animals \citep{wann19,renner18,chu17}. % emwJun4 -- check last three refs here are still relevant

@@

% emwJun4 -- check below where I added 'spring phenology' -- make sure that's accurate (I also removed the decimals)
Evidence for local adaptation in spring phenology comes mostly from common garden studies. For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through these experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}. Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (For example,\emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}). This has led to some debate over the prevalence and importance of adaptation in spring tree phenology. While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they appear potentially more common in European studies. % emwJun4 -- we need to justify why we say they are more common in Europe (I know I wrote this! But you should change anything I write that seems incorrect to you and let me know) ... if you can, please add an explanation and citations with the explanation (or we change 'they appear potentially more common in European studies,' with 'they could vary by continent').

@@

 %REFS
 % savolainenAREES -- have not read recently but sounds relevant
 % gauzere2020 -- cite and read
 % \cite{chang2021} and \cite{chuine2016} -- relevant for discussion of chilling
 % See also Salk 2020 paper and some papers mentioned above

@@

\section{Discussion}
%emw2Mar -- there's a delicate balance of how much rationale you can give for your methods in the methods (and similar for the results). I would move the below to the discussion section. 
While coarse metrics such as latitude and MAT may represent how similar the climates are between the provenances and gardens in times that matter for the events. If climates are very similar, then we would expect similar timings [add more here].

@@

The weak relationship between spring event dates and provenance latitude and MAT that we find in European studies might be explained by the higher extent of climate overlap in those studies. The more similar the climate is between provenances and gardens, the less difference between spring event dates.

The inconsistent and weak clines in spring events that we found suggest high plasticity in spring phenology across continents and species. Fall events, on the other hand, exhibit stronger clines which suggest more local adaptation, especially in North America. Overall, our results predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.

In contrast to spring events, we found strong latitudinal clines in fall events across both continents, with local adaptation appearing much stronger in North America than in Europe. Our results show that spring events are highly plastic, and thus may shift with warming, but data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, are needed for forecasting.

\begin{enumerate}
\item Opening: 1-2 paragraph overview of our findings (spring phenology show no real clines, some variation by continent and gymnosperm/angiosperm but hard to fully disentangle). Suggests some small clines of spring phenology in some studies, but no consistent effects 
% \item Implications of our results on animals: Many terrestrial animals time their breeding season to coincide with the availability of food resources. For example, migratory birds arrive at breeding grounds amidst the peaking of insects \citep{hallfors20}. Many butterfly species ensure their larvae have an ample food supply by synchronizing their egg-laying with the phenology of their host plants \citep{rytteri21,posle18}. Similarly, the migration and spawning of many marine organisms are closely tied to water temperature and the availability of planktonic food sources \citep{genner10}. Such synchronization enables them to provide optimal food for their young during critical growth stages. Therefore, climate warming and the associated changes in spring phenology timing, such as earlier leafout or insect emergence due to climate change, can disrupt this synchronization and negatively impact animal populations \citep{wann19,renner18, chu17}. 
 \item Subheader: Adaptation in spring vs fall phenology: Implications for climate change responses
 \begin{enumerate}
 \item Mostly unorganized list of things we may want to say in this section
 \begin{enumerate}
 \item Current advances in spring phenology may continue, but delays in fall phenology will depend on speed of adaptation
 \item If current advances in spring phenology continues it could mean shifts in spring frost etc.
  \item Spring phenology, specifically the timing of budburst and leaf flush, plays a significant role in determining plant fitness and has been widely adopted as a useful tool to investigate plant adaptation mechanisms \citep{guo22, chuine01}. In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}. Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}. Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}.
 \item Clines in fall phenology are stronger and more consistent while clines in spring phenology are weaker and can go either way  (highlight that some studies and we should include WHICH ones do show clines by XX days).
 \item Mention some of of the co and counter-gradient studies here
  \item  An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
 \item Implications of small differences in spring phenology could mean different gene flow etc. 
  \item the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}. 
  \item Need to cite elevation studies and discuss them somewhere here.
  \item Somewhere mention results from chamber studies and local adaptation (which are hard to interpret given issues with chilling) 
 \item What else?
 \end{enumerate}
  \end{enumerate}
 \item Subheader: Variation across continents and species groupings
 \begin{enumerate}
 \item  Unorganized list of things we may want to say in this section
 \begin{enumerate}
 \item No strong evidence of continental differences, could make sense based on long-term climate history perhaps ... 
\item discuss xylem (ring versus diffuse porous spp) and embolism but might be too in the weeds (see Salk article)
\item Angiosperm vs gymnosperm
 \end{enumerate}
 \end{enumerate}
\end{enumerate}

% Must cover in discussion:

% Counter vs. co gradient literature
% Maybe also discuss? ice sheet dynamics in Europe/North America, which could impact rates of local adaptation.

>>>>

% emwJun4: I really love the info you added here, but we have to be careful to not focus too deeply on synchrony as it is a very specific topic and our work is broader than that (sorry if I confused you by my request for 'refs' -- I mostly wanted just citations; I suspect we can use some of the text for the discussion so it will not go to waste). We also need to keep paragraphs themselves and our intro overall short, so I cut here. Thus I slimmed this way down and kept many references, but less of the examples.
Many of the currently observed responses to climate change appear to be mostly plastic \citep{burton22,zettle21,bonamour19, king17}, including the most reported biological response to climate change---shifting phenology. Phenology (the timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and exploit the resources of each growing season \citep{chuine10,hanninen11,rytteri21,posle18}. As such, phenology plays a significant role in determining fitness for both plants \citep{guo22,chuine01} and animals \citep{wann19,renner18,chu17}.

@@

Evidence for local adaptation in spring phenology comes mostly from common garden studies. For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through these experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}. Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (For example,\emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}). This has led to some debate over the prevalence and importance of adaptation in spring tree phenology. While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they could vary by continent.

@@

@@

\section{Discussion}

\begin{enumerate}
\item Opening: 1-2 paragraph overview of our findings (spring phenology show no real clines, some variation by continent and gymnosperm/angiosperm but hard to fully disentangle). Suggests some small clines of spring phenology in some studies, but no consistent effects 
% The inconsistent and weak clines in spring events that we found suggest high plasticity in spring phenology across continents and species. Fall events, on the other hand, exhibit stronger clines which suggest more local adaptation, especially in North America. Overall, our results predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures. The weak relationship between spring event dates and provenance latitude and MAT that we find in European studies might be explained by the higher extent of climate overlap in those studies. The more similar the climate is between provenances and gardens, the less difference between spring event dates. In contrast to spring events, we found strong latitudinal clines in fall events across both continents, with local adaptation appearing much stronger in North America than in Europe. Our results show that spring events are highly plastic, and thus may shift with warming, but data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, are needed for forecasting.

% \item Implications of our results on animals: Many terrestrial animals time their breeding season to coincide with the availability of food resources. For example, migratory birds arrive at breeding grounds amidst the peaking of insects \citep{hallfors20}. Many butterfly species ensure their larvae have an ample food supply by synchronizing their egg-laying with the phenology of their host plants \citep{rytteri21,posle18}. Similarly, the migration and spawning of many marine organisms are closely tied to water temperature and the availability of planktonic food sources \citep{genner10}. Such synchronization enables them to provide optimal food for their young during critical growth stages. Therefore, climate warming and the associated changes in spring phenology timing, such as earlier leafout or insect emergence due to climate change, can disrupt this synchronization and negatively impact animal populations \citep{wann19,renner18, chu17}. 
 \item Subheader: Adaptation in spring vs fall phenology: Implications for climate change responses
 \begin{enumerate}
 \item Mostly unorganized list of things we may want to say in this section
 \begin{enumerate}
 \item Current advances in spring phenology may continue, but delays in fall phenology will depend on speed of adaptation
 \item If current advances in spring phenology continues it could mean shifts in spring frost etc.
  \item Spring phenology, specifically the timing of budburst and leaf flush, plays a significant role in determining plant fitness and has been widely adopted as a useful tool to investigate plant adaptation mechanisms \citep{guo22, chuine01}. In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}. Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}. Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}.
 \item Clines in fall phenology are stronger and more consistent while clines in spring phenology are weaker and can go either way  (highlight that some studies and we should include WHICH ones do show clines by XX days).
 \item Mention some of of the co and counter-gradient studies here
  \item  An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
 \item Implications of small differences in spring phenology could mean different gene flow etc. 
  \item the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}. 
  \item Need to cite elevation studies and discuss them somewhere here.
  \item Somewhere mention results from chamber studies and local adaptation (which are hard to interpret given issues with chilling) 
 \item What else?
 % AlinaJune10: Maybe also discuss ice sheet dynamics in Europe/North America, which could impact rates of local adaptation.
 \end{enumerate}
  \end{enumerate}
 \item Subheader: Variation across continents and species groupings
 \begin{enumerate}
 \item  Unorganized list of things we may want to say in this section
 \begin{enumerate}
 \item No strong evidence of continental differences, could make sense based on long-term climate history perhaps ... 
\item discuss xylem (ring versus diffuse porous spp) and embolism but might be too in the weeds (see Salk article)
\item Angiosperm vs gymnosperm
 \end{enumerate}
 \end{enumerate}
\end{enumerate}

%emw2Mar -- there's a delicate balance of how much rationale you can give for your methods in the methods (and similar for the results). I would move the below to the discussion section. 
While coarse metrics such as latitude and MAT may represent how similar the climates are between the provenances and gardens in times that matter for the events. If climates are very similar, then we would expect similar timings [add more here].

@@

 %REFS
 % savolainenAREES -- have not read recently but sounds relevant
 % gauzere2020 -- cite and read
 % \cite{chang2021} and \cite{chuine2016} -- relevant for discussion of chilling
 % See also Salk 2020 paper

### 2023-06-13 Docs/manuscript.tex
edited intro/discussion

Many of the currently observed responses to climate change appear to be mostly plastic \citep{burton22,zettle21,bonamour19, king17}, including the most reported biological response to climate change---shifting phenology. Phenology (timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and fully exploit the resources of each growing season \citep{chuine10,hanninen11}. Many terrestrial animals time their breeding season to coincide with the availability of food resources. For example, migratory birds arrive at breeding grounds amidst the peaking of insects \citep{hallfors20}. Many butterfly species ensure their larvae have an ample food supply by synchronizing their egg-laying with the phenology of their host plants \citep{rytteri21,posle18}. Similarly, the migration and spawning of many marine organisms are closely tied to water temperature and the availability of planktonic food sources \citep{genner10}. Such synchronization enables them to provide optimal food for their young during critical growth stages. Therefore, climate warming and the associated changes in spring phenology timing, such as earlier leafout or insect emergence due to climate change, can disrupt this synchronization and negatively impact animal populations \citep{wann19,renner18, chu17}. As such, phenology plays a significant role in determining fitness \citep{guo22, chuine01}. Shifted phenology in recent decades---with many events moving several days per decade \citep{vita21,khar18,Menzel06} ---has thus led to concerns about fitness consequences, and the limits of possible future shifts. Future phenological shifts will depend on how much phenology is determined by adaptation versus plasticity will determine, yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.

Tree phenology is critically important to climate change forecasting and has been an area of long-term study of adaptation and plasticity \citep{Cleland:2007or}. The timing of budburst and senescence can impact plant competition \citep{fridley12}, growth \citep{myneni97}, and carbon sequestration \citep{Barichivich12}---thus affecting climate change itself. Hundreds of years of study have established common clines in fall phenology, as locations with shorter growing season (poleward, higher elevation) exhibit earlier stopping of growth (e.g., via budset), and connected these to an underlying proximate mechanism of changing photoperiod cues, driven by local adaptation to the local growing season and relevant photoperiod to stop growth \citep{Alberto13,Savolainen07}. In contrast, spring phenology appears far more plastic \citep{AitkenBemmels16} and determined more strongly by temperature \citep{flynn18}. Many studies, however, have argued that spring phenology shows important levels of adaptation.

Evidence for local adaptation in spring phenology comes mostly from common garden studies. For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through these experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}. Studies have documented provenance differences of 2.08-4.17 days per degree latitude for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (For example,\emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}). This has led to some debate over the prevalence and importance of adaptation in spring tree phenology. While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they appear potentially more common in European studies.

Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, as North American springs are more variable across years than European ones \citep{tward21,zohner2017,schwartz00}. Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years. Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{cat23, ospreebbms,ospreephoto}, but are poorly controlled compared to traditional provenance trials, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}. Even for more carefully designed common garden differences in study design---especially studied species---may complicate understanding what underlies potential trends across continents, with European studies focusing more on deciduous angiosperms and North American studies focusing often on evergreen gymnosperms.  
To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring and fall events. We test for evidence of adaptation via provenance trends with latitude and climate. Further, to better understand these clines, we also examine the following:
\begin{itemize}
\item How different are clines between spring and fall phenology? 
\item How different are clines between Europe and North America?
\item How different are clines between angiosperm and gymnosperm species?
\end{itemize}
To address these questions, we combined a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology with Bayesian hierarchical models.

 %REFS
 % savolainenAREES -- have not read recently but sounds relevant
 % gauzere2020 -- cite and read
 % \cite{chang2021} and \cite{chuine2016} -- relevant for discussion of chilling
 % See also Salk 2020 paper and some papers mentioned above

@@

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.jpg}
    \caption{Event day of year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models. Spring events shown on top and fall events at the bottom. European Gardens and species are bolded and denoted by an asterisk (*). } 
    \label{figure:springfall_latmat}
\end{figure}

In contrast, fall events (budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (3.16 days/degree [2.87-3.45]) and the decrease in MAT (4.78 days/$^{\circ}$C [4.1 - 5.4]), meaning fall events WEre earlier where provenance MAT was lower (Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continental_spptype_effect}A).

@@

Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
Spring events slightly delayed 0.37 (0.15 - 0.59) days per degree north in angiosperms and 0.23 (0.00 - 0.46) days per degree north in gymnosperms. Fall events advanced 3.18 (2.76 - 3.62) days per degree north in angiosperms and 3.14 (2.81-3.47) days per degree north in gymnosperms.
However, effects of MAT on spring events weakly diverged (Fig.\ref{figure:continental_spptype_effect}B). Spring events advanced 0.82 (0.54 - 1.11) days/$^{\circ}$C as MAT increased in angiosperms and delayed 0.76 (0.37 - 1.14) days/$^{\circ}$C as MAT increased in gymnosperms. Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (6.23 days) than angiosperms (3.69 days) (Fig. \ref{figure:continental_spptype_effect}B).

%EMWApr20 -- need to refer to some supp figures or tables or such below. Once you have the supp so I can see better what we're referring to here in first sentence we should likely edit this paragraph a little. 
While we believe that coarse metrics such as latitude and MAT would ultimately represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months that matter for the events to further test how much climate similarity between provenances and gardens predicts local adaptation. For spring events, we considered overlap across March to May. However, results were not qualitatively different than using MAT (See Fig. \ref{supplement-figure:overlap_scatterplot} in Supplements). We observed very weak effects of climate overlap on spring events (0.01 [0.02 - 0.03] days per one per cent increase in climate overlap), nearly identical across angiosperms (0.02 [0.00 - 0.05]) and gymnosperms (0.04 [0.00 - 0.09]). Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing 0.72 [0.51 - 0.92] days per one per cent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).

@@

The weak relationship between spring event dates and provenance latitude and MAT that we find in European studies might be explained by the higher extent of climate overlap in those studies. The more similar the climate is between provenances and gardens, the less difference between spring event dates.

The inconsistent and weak clines in spring events that we found suggest high plasticity in spring phenology across continents and species. Fall events, on the other hand, exhibit stronger clines which suggest more local adaptation, especially in North America. Overall, our results predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.

In contrast to spring events, we found strong latitudinal clines in fall events across both continents, with local adaptation appearing much stronger in North America than in Europe. Our results show that spring events are highly plastic, and thus may shift with warming, but data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, are needed for forecasting.

\begin{enumerate}
\item Opening: 1-2 paragraph overview of our findings (spring phenology show no real clines, some variation by continent and gymnosperm/angiosperm but hard to fully disentangle). Suggests some small clines of spring phenology in some studies, but no consistent effects
 \item Subheader: Adaptation in spring vs fall phenology: Implications for climate change responses
 \begin{enumerate}
 \item Mostly unorganized list of things we may want to say in this section
 \begin{enumerate}
 \item Current advances in spring phenology may continue, but delays in fall phenology will depend on speed of adaptation
 \item If current advances in spring phenology continues it could mean shifts in spring frost etc.
  \item Spring phenology, specifically the timing of budburst and leaf flush, plays a significant role in determining plant fitness and has been widely adopted as a useful tool to investigate plant adaptation mechanisms \citep{guo22, chuine01}. In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}. Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}. Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}.
 \item Clines in fall phenology are stronger and more consistent while clines in spring phenology are weaker and can go either way  (highlight that some studies and we should include WHICH ones do show clines by XX days).
 \item Mention some of of the co and counter-gradient studies here
  \item  An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
 \item Implications of small differences in spring phenology could mean different gene flow etc. 
  \item the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}. 
  \item Need to cite elevation studies and discuss them somewhere here.
  \item Somewhere mention results from chamber studies and local adaptation (which are hard to interpret given issues with chilling) 
 \item What else?
 \end{enumerate}
  \end{enumerate}
 \item Subheader: Variation across continents and species groupings
 \begin{enumerate}
 \item  Unorganized list of things we may want to say in this section
 \begin{enumerate}
 \item No strong evidence of continental differences, could make sense based on long-term climate history perhaps ... 
\item discuss xylem (ring versus diffuse porous spp) and embolism but might be too in the weeds (see Salk article)
 \end{enumerate}
 \end{enumerate}
\end{enumerate}

% Must cover in discussion:

% Counter vs. co gradient literature
% Maybe also discuss? ice sheet dynamics in Europe/North America, which could impact rates of local adaptation.

>>>>

% emwJun4: I really love the info you added here, but we have to be careful to not focus too deeply on synchrony as it is a very specific topic and our work is broader than that (sorry if I confused you by my request for 'refs' -- I mostly wanted just citations; I suspect we can use some of the text for the discussion so it will not go to waste). We also need to keep paragraphs themselves and our intro overall short, so I cut here. Thus I slimmed this way down and kept many references, but less of the examples.
Many of the currently observed responses to climate change appear to be mostly plastic \citep{burton22,zettle21,bonamour19, king17}, including the most reported biological response to climate change---shifting phenology. Phenology (the timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and exploit the resources of each growing season \citep{chuine10,hanninen11,rytteri21,posle18}. As such, phenology plays a significant role in determining fitness for both plants \citep{guo22,chuine01} and animals \citep{wann19,renner18,chu17}.

Shifted phenology in recent decades---with many events moving several days per decade \citep{vita21,khar18,Menzel06}---has led to concerns about fitness consequences, and the limits of possible future shifts. Future phenological shifts will depend on how much phenology is determined by adaptation versus plasticity; yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.

Tree phenology is critically important to climate change forecasting and has been an area of long-term study of adaptation and plasticity \citep{Cleland:2007or}. The timing of budburst and senescence can impact plant competition \citep{fridley12}, growth \citep{myneni97}, and carbon sequestration \citep{Barichivich12}---thus affecting climate change itself. Hundreds of years of study have established common clines in fall phenology, as locations with shorter growing season (poleward and higher elevations) exhibit earlier stopping of growth (e.g., budset). Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold that trigger budset), driven by local adaptation to the local growing season \citep{Alberto13,Savolainen07}. In contrast, spring phenology appears far more plastic \citep{AitkenBemmels16} and determined more strongly by temperature \citep{flynn18}. Many studies, however, have argued that spring phenology shows important levels of adaptation.

Evidence for local adaptation in spring phenology comes mostly from common garden studies. For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through these experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}. Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (For example,\emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}). This has led to some debate over the prevalence and importance of adaptation in spring tree phenology. While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they could vary by continent.

Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, as North American springs are more variable across years than European ones \citep{tward21,zohner2017,schwartz00}. Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years. Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{zohner2017}, but are poorly controlled compared to traditional provenance trials, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}. Even for more carefully designed common garden differences in study design---especially studied species---may complicate understanding what underlies potential trends across continents, with European studies focusing more on deciduous angiosperms and North American studies focusing often on evergreen gymnosperms.

To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring and fall events. We test for evidence of adaptation via provenance trends with latitude and climate. Further, to better understand these clines, we also examine the following:
\begin{itemize}
\item How different are clines between spring and fall phenology? 
\item How different are clines between Europe and North America?
\item How different are clines between angiosperm and gymnosperm species?
\end{itemize}
To address these questions, we combined Bayesian hierarchical models with a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology.

@@

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.jpg}
    \caption{Event day of year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models. Spring events shown on top and fall events at the bottom. European gardens and species are bolded and denoted by an asterisk (*). } 
    \label{figure:springfall_latmat}
\end{figure}

In contrast, fall events (budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (3.16 days/degree [2.87-3.45]) and the decrease in MAT (4.78 days/$^{\circ}$C [4.1 - 5.4]), meaning fall events Were earlier where provenance MAT was lower (Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig. \ref{figure:continental_spptype_effect}A).

@@

% emwJun4 -- can you reference the figures in the supp that have the latitude results?
Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
Spring events slightly delayed 0.37 (0.15 - 0.59) days per degree north in angiosperms and 0.23 (0.00 - 0.46) days per degree north in gymnosperms. Fall events advanced 3.18 (2.76 - 3.62) days per degree north in angiosperms and 3.14 (2.81-3.47) days per degree north in gymnosperms.
However, effects of MAT on spring events weakly diverged (Fig.\ref{figure:continental_spptype_effect}B). Spring events advanced 0.82 (0.54 - 1.11) days/$^{\circ}$C as MAT increased in angiosperms and delayed 0.76 (0.37 - 1.14) days/$^{\circ}$C as MAT increased in gymnosperms. Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (6.23 days) than angiosperms (3.69 days) (Fig. \ref{figure:continental_spptype_effect}B).

%EMWApr20 -- need to refer to some supp figures or tables or such below. Once you have the supp so I can see better what we're referring to here in first sentence we should likely edit this paragraph a little. 
% emwJun4 -- give more explanation of what we mean by  'in months that matter for the events' ... is this in the methods? 
While we believe that coarse metrics such as latitude and MAT would ultimately represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months that matter for the events to further test how much climate similarity between provenances and gardens predicts local adaptation. For spring events, we considered overlap across March to May. However, results were not qualitatively different than using MAT (See Fig. \ref{supplement-figure:overlap_scatterplot} in Supplements). We observed very weak effects of climate overlap on spring events (0.01 [0.02 - 0.03] days per one per cent increase in climate overlap), nearly identical across angiosperms (0.02 [0.00 - 0.05]) and gymnosperms (0.04 [0.00 - 0.09]). Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing 0.72 [0.51 - 0.92] days per one per cent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).

@@

 %REFS
 % savolainenAREES -- have not read recently but sounds relevant
 % gauzere2020 -- cite and read
 % \cite{chang2021} and \cite{chuine2016} -- relevant for discussion of chilling
 % See also Salk 2020 paper
